Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka

### 2025-08-15 analysis/paper.qmd
Finalise resubmission

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 82 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  <!-- TODO: rewrite following reviews. Highlight BA to EH decline vs. 'amelioration' concept -->
  The models predict significant reductions and/or shifts in species ranges in
  the terminal Pleistocene and Early Holocene compared to present conditions. 
  Many species that are found throughout the region's 'hilly flanks' today 
  are indicated to have much more restricted distributions centered on the 
  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
  The models performance in predicting the occurrence of specific species at 
  archaeological sites is highly variable, but in aggregate the predictions are
  coherent and align with broad-scale trends in the ubiquity and relative 
  abundance of species in the archaeological record.<!--
  however [...]
  The regional ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.-->
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
  cache: true
---

<!--
TODO before publication:
- [ ] Revise discussion of archaeo verification
- [ ] Outstanding TODOs
- [ ] Add missing references
- [x] Appendix with all hindcast predictions (HTML?)
-   [] TIFs?
- [ ] Clean up package code
- [ ] Zenodo release
- [ ] Final proofread
-->

```{r dependencies, cache=FALSE, message=FALSE}
library("cowplot")
library("dplyr")
library("dotenv")
library("english")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("probably")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

In this paper we present a complementary, deductive approach based on ecological niche modelling.
Rather than inferring environmental conditions from preserved physical evidence, 
we predict the ranges of individual species relevant to human subsistence based on a model of their current environmental niche and simulations of past palaeoclimate. 
Though hypothetical, this gives us an independent line of evidence on past ecologies that is independent of the environmental archaeological and palaeoclimatic records.
This means that the ancient data can be reserved for assessing the model's ability to 'hindcast' past conditions.
In this sense, discrepencies between the two records are perhaps the most interesting result, as they indicate processes affecting one or both records that are not fully accounted for and therefore generate new questions.
Our computational approach is also readily scaled up, allowing us to model spatially-explicit palaeodistributions for a large number of species, for the whole region, under multiple past climatologies.

@@

<!-- TODO: figure highlighting discrepencies between 'expert interpolation' -->

@@

Ecological niche modelling or species distribution modelling is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors [@FranklinMiller2009].
Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
This model can then be used to predict the range of the organism in question either in the same or a different environment.
@TownsendPetersonSoberon2012 suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.

Ecological niche modelling has long been of interest to archaeologists as both a means of exploring the biological niche of humans and for reconstructing the past environments they inhabited [@DavidPollyEronen2011; @FranklinEtAl2015].
In the first sense, it has been used most extensively to model the range of humans and other hominin species [e.g. @BenitoEtAl2017; @YousefiEtAl2020; @BanksEtAl2021; @YaworskyEtAl2024a; @YaworskyEtAl2024b; @GuranEtAl2024], especially in the Palaeolithic.
This overlaps with what archaeologists usually call generically 'predictive modelling' [@VerhagenWhitley2020]—or more precisely 'site distribution modelling'—which is essentially the same approach as (and often borrows methodologies from) ecological niche modelling but applied to the occurrence of archaeological sites.
Here what is modelled is not strictly a biological niche alone, but also aspects of human geography, taphonomy, and archaeological visibility.
These applications can be distinguished from 'palaeoecological niche modelling', where the object of model remains, as in ecology, a non-human biological niche.

@@

ne_land <- read_sf(raw_data("ne"), "ne_10m_land") |> 
  st_geometry() |>
  st_crop(buffer_bbox(w_eurasia, 10)) |>
  st_difference(ne_ocean) |>
  # st_difference(ne_lakes) |>
  st_union()

@@

@@

      # TODO: this creates duplicate rows

@@

```{r fig-region}
#| fig-cap: Late Epipalaeolithic & Pre-Pottery Neolithic archaeological sites used to generate modelled flora
# TODO: Update to show Eurasia (with zoom-in for sites?)
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
  layer_spatial(w_asia, fill = NA) +
  layer_spatial(archaeo_sites, size = 0.5) +
  scale_linewidth_identity() +
  coord_sf(crs = w_asia_albers)
```

@@

<!-- TODO generate GBIF occurrence citation:
https://docs.ropensci.org/rgbif/articles/getting_occurrence_data.html -->

```{r data-occ}
# Download occurrence data from GBIF, generating a citable derived dataset
if (length(dir_ls(derived_data("gbif"))) == 0) {
  # Requires GBIF credentials as environment variables, see:
  # https://docs.ropensci.org/rgbif/articles/gbif_credentials.html
  message("Reacquiring GBIF occurrences...")
  gbif_download <- occ_download(
    pred_in("taxonKey", flora$gbif_key),
    pred_within(wk::as_wkt(w_eurasia)),
    pred_default(), # Filter common data quality issues, see ?rgbif::pred_default() for details
    # pred_lt("coordinateUncertaintyInMeters", 1000), # to strict for W. Asia!
    format = "SIMPLE_CSV"
  ) |>
    occ_download_wait() # wait for dataset to be prepared

@@

flora_occ_meta <- read_csv(dir_ls(derived_data("gbif"), glob = "*.csv")[1])

@@

<!-- TODO: cite dataset -->
Georeferenced occurrence data for West Eurasia between 0 and 60° of latitude was obtained from the Global Biodiversity Information Facility (GBIF) using via and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
The GBIF dataset [@GBIFDerivedData] excluded fossil occurrences, recorded absences, and records with missing or dubious coordinates.
Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we did not attempt to model `r english(nrow(filter(flora_all, n_occ < min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
Multiple records of the same taxon at the same coordinate were discarded because they do not impart information to the model.
The resulting cleaned dataset used to train our niche models comprises `r flora_occ_meta$totalRecords` from `r flora_occ_meta$numberDatasets` constituent datasets.

```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia used to train models (N=", nrow(unnest(flora, occ)))}
# TODO: Update, Eurasia
flora |>
 unnest(occ) |>
 distinct(latitude, longitude) |>
 st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
 ggplot() +
 annotation_spatial(ne_land, fill = "#ffffff") +
 geom_sf(shape = ".") +
 layer_spatial(w_asia, fill = NA) +
 coord_sf(crs = w_asia_albers, default_crs = latlong)
```

GBIF is currently the best available general-purpose occurrence dataset for the West Asia region, its coverage is uneven both geographically and from species to species.
@fig-occ-map shows clearly that the Southern Levant, and Israel specifically, is significantly more densely sampled than other parts of West Asia.

```{r data-background}
# TODO: slow - cache
if (!file_exists(derived_data("flora_background.Rdata"))) {
flora <- flora |>
  mutate(
    occ = map(occ, mutate, present = TRUE, weight = 1),
    bg = map(occ, function(x, region, n) { 
      background_sample(
        region = region, 
        N = n,
        coord_x = "longitude", coord_y = "latitude",
        present = FALSE,
        weight = (1 / n) * nrow(x) # weight presences and background equally
      )
    }, region = w_eurasia_land, n = n_background),
    occ = map2(occ, bg, bind_rows)
  ) |>
  select(-bg)
  saveRDS(flora, derived_data("flora_background.Rdata"))
}
flora <- readRDS(derived_data("flora_background.Rdata"))
```

@@

```{r data-climate}
# Reacquire paleoclim data if needed:
if (!file_exists(derived_data("climate.nc"))) {
  if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
    message("Redownloading Paleoclim data...")
    source(here("analysis", "data", "data-paleoclim.R"))
  }

@@

climate <- read_stars(derived_data("climate.nc"), proxy = TRUE)
# Recover dimension values lost in NC conversion
climate <- st_set_dimensions(climate, "period", values = climate_periods$code)
```

@@

```{r data-flora-predictors}
# TODO: factor out
bind_extract <- function(x, y) {
  extract <- st_extract(y, st_transform(x, st_crs(y)))
  cbind(x, select(tibble(extract), -geometry))
}

# TODO: slow, cache
flora <- flora |> 
  mutate(
    occ = map(occ, st_as_sf, coords = c("longitude", "latitude"), crs = latlong, 
              remove = FALSE),
    occ = map(occ, bind_extract, st_as_stars(hack(climate, "period", "cur"))),
    occ = map(occ, bind_extract, st_as_stars(terrain)),
    occ = map(occ, bind_extract, st_as_stars(soil))
  )
```

@@

```{r niche-data}
# Split training and test sets for model assessment 
flora <-  flora |>
  # Manually impute some problematic cases
  # mutate(
  #   occ = case_when(
  #     taxon_key %in% c(12225945, 7405743) ~ map(occ, function(x) {
  #       mutate(x, across())
  #     })
  #   )
  # ) |>
  mutate(
    # De-sf
    occ = map(occ, \(x) select(tibble(x), -geometry)),
    # TODO: can this happen earlier?
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE)))),
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
  )

@@

```{r niche-predict, cache.lazy=FALSE}
# NOTE: high memory usage
if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
  flora <- flora |>
    mutate(
      pred = map(niche, collect_predictions),
      palaeodist = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
                       obj = env_w_asia, .progress = TRUE),
      # Binarize
      palaeodist = map(palaeodist, mutate, present = prediction.TRUE. > 0.5),
    )
  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))

# TODO write TIFs
```

@@

We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition [@tbl-results-summary].
Model accuracy (proportion of correctly classified presence and background samples) ranged between `r percent(filter(niche_test_summary, .metric == "accuracy")$min)` and `r percent(filter(niche_test_summary, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "accuracy")$mean)`.
Sensitivity (proportion of correctly classified presence samples) ranged between `r percent(filter(niche_test_summary, .metric == "sens")$min)` and `r percent(filter(niche_test_summary, .metric == "sens")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "sens")$mean)`.
The area under the models' receiver operating characteristic curves (ROC-AUC) was on average `r round(filter(niche_test_summary, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_summary, .metric == "accuracy")$sd, 3)`
Model sensitivity is loosely correlated with the number of occurrences available for training (@fig-model-sensitivity-vs-n_occ), with the worst-performing models all having less than 300 recorded occurrences: `r knitr::combine_words(paste0("*", filter(unnest(flora, niche_test_metrics), .metric == "sens" & .estimate < 0.85)$taxon, "*"))`.
Test metrics and ROC curves for the individual models are included in the appendix.

```{r niche-predict-archaeo, cache.lazy=FALSE}
# Extract predictors and make predictions for archaeobotanical occurrences
# TODO: what about negatives?
if (!file.exists(derived_data("archaeo_pred.Rdata"))) {
  archaeo_pred <- flora |>
    unnest(archaeo) |>
    # Add dummy GBIF id
    mutate(gbif_id = 0) |>
    nest(.by = c(climate_period), .key = "archaeo") |>
    mutate(
      archaeo = map(archaeo, st_as_sf, coords = c("longitude", "latitude"),
                    crs = latlong, remove = FALSE),
      archaeo = map2(archaeo, climate_period, function(x, period) {
        bind_extract(x, st_as_stars(hack(climate, "period", as.character(period))))
      }),
      archaeo = map(archaeo, bind_extract, st_as_stars(terrain)),
      archaeo = map(archaeo, bind_extract, st_as_stars(soil))
    ) |>
    unnest(archaeo) |>
    nest(.by = any_of(colnames(flora)), .key = "archaeo") |> # Restore original nesting
    mutate(
      archaeo_pred = map2(niche, archaeo, function(x, y) {
        # TODO: shouldn't be necessary - remove from workflow
        y <- mutate(y, gbif_key = NA, scientific_name = NA, species = NA)
        augment(extract_workflow(x), new_data = y)
      })
    ) |>
    select(taxon, archaeo_pred)

@@

The ability of the hindcast models to predict the occurrence of specific species at archaeological sites is worse, with `r percent(p_archaeo_pred_correct)` of presences in archaeobotanical assemblages successfully predicted.
Model sensitivity (proportion of corrected predicted presenses) in relation to the archaeological data is on average `r round(mean(archaeo_pred_summary$p_predicted, na.rm=TRUE), 2)`±`r round(sd(archaeo_pred_summary$p_predicted, na.rm=TRUE), 2)` [@tbl-results-summary].
A full assessment of the hindcasting performance of the individual models can be found in the appendix.
<!-- TODO: negative -->

@@

# TODO:
# - alt. taxons as footnotes
# - check boxes of which (climate?) periods observed in ?
results_summary <- flora_all |>
  left_join(flora, join_by(taxon, taxon_aliases, archaeo, n_archaeo)) |>
  left_join(archaeo_pred_summary, join_by(taxon)) |>
  mutate(n_occ = coalesce(n_occ.y, n_occ.x), .before = n_occ.x) |>
  # Extract model metrics
  unnest(c(niche_test_metrics), keep_empty = TRUE) |>
  select(-.estimator, -.config) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  # Excluded samples
  mutate(exclude = (n_occ < min_n_occ) & 
           !taxon %in% c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")) |>
  # Prepare table
  select(taxon, taxon_aliases, n_archaeo, n_occ, exclude, accuracy, roc_auc, 
         sens, p_predicted) |>
  arrange(taxon)

@@

## Reduction in range sizes over the Pleistocene/Holocene boundary

@@

* Time-averaged, sometimes quite severely
* (Understudied) possibility for human transplanting and transport; we know this is the dawn of agriculture!
* Generally small sample sizes, so individual misidentifications have a large impact

So what can we do about it???

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: jar@hum.ku.dk
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 65 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential niches of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  Our models performed well against independent contemporary test data, but 
  their ability to predict the occurrence of specific species at 
  archaeological sites was much more variable, probably reflecting a tendency of 
  the method to underestimate the species' fundamental niche. Nevertheless, the 
  majority of species are predicted to have had more restricted geographic 
  distributions under past climate conditions compared to today. Crop 
  progenitors and several wild food species modelled are modelled to have been 
  concentrated in the Levant and, to a lesser extent, Cyprus and Western 
  Anatolia, suggesting these regions may have served as glacial *refugia*. 
  The average size of species' niche shrunk by an average of c. 25% from the 
  terminal Pleistocene to the Early Holocene, indicating that economically
  significant plants were adapted to cryo-arid conditions and did not, as often
  assumed, initially respond positively to the 'ameliorated' climate of the 
  Holocene.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
  docx:
    toc: true
    number-sections: true
bibliography: 
  - references.bib
  - gbif.bib
execute: 
  echo: false
  cache: true
---

<!--
TODO before publication:
- [ ] Zenodo release
- [ ] Clean up package code
-->

```{r dependencies, cache=FALSE, message=FALSE}
library("cowplot")
library("dplyr")
library("dotenv")
library("english")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("patchwork")
library("probably")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

In this paper we present a complementary, deductive approach based on ecological niche modelling.
Rather than inferring environmental conditions from preserved physical evidence, 
we predict the potential niche of individual species relevant to human subsistence based on a model of their current environmental niche and simulations of past palaeoclimate. 
Though hypothetical, this gives us an independent line of evidence on past ecologies that is independent of the environmental archaeological and palaeoclimatic records.
This means that the ancient data can be reserved for assessing the model's ability to 'hindcast' past conditions.
In this sense, discrepencies between the two records are perhaps the most interesting result, as they indicate processes affecting one or both records that are not fully accounted for and therefore generate new questions.
Our computational approach is also readily scaled up, allowing us to model spatially-explicit palaeodistributions for a large number of species, for the whole region, under multiple reconstructed climate scenarios.

@@

@@

Ecological niche modelling or species distribution modelling is widely used by ecologists to mathematically predict the geographic distribution of a species from a sample of occurrences [@FranklinMiller2009; @SilleroEtAl2021].
Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
This model can then be used to predict the potential niche of the organism in question either in the same or a different environment.
@TownsendPetersonSoberon2012 suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past niches.
Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.

Ecological niche modelling has long been of interest to archaeologists as both a means of exploring the biological niche of humans and for reconstructing the past environments they inhabited [@DavidPollyEronen2011; @FranklinEtAl2015].
In the first sense, it has been used most extensively to model the niche of humans and other hominin species [e.g. @BenitoEtAl2017; @YousefiEtAl2020; @BanksEtAl2021; @YaworskyEtAl2024a; @YaworskyEtAl2024b; @GuranEtAl2024], especially in the Palaeolithic.
This overlaps with what archaeologists usually call generically 'predictive modelling' [@VerhagenWhitley2020]—or more precisely 'site distribution modelling'—which is essentially the same approach as (and often borrows methodologies from) ecological niche modelling but applied to the occurrence of archaeological sites.
Here what is modelled is not strictly a biological niche alone, but also aspects of human geography, taphonomy, and archaeological visibility.
These applications can be distinguished from 'palaeoecological niche modelling', where the object of model remains, as in ecology, a non-human biological niche.

@@

ne_land <- read_sf(raw_data("ne"), "ne_10m_land") |> 
  st_geometry() |>
#  st_crop(buffer_bbox(w_eurasia, 10)) |>
  st_difference(ne_ocean) |>
  # st_difference(ne_lakes) |>
  st_union()

@@

    filter(taxon != "Medicago astroites") |> # Not reliably identifiable

@@

@@

```{r fig-region}
#| fig-cap: Map of the study region (West Asia, grey box) with locations of Late Epipalaeolithic and Pre-Pottery Neolithic archaeobotanical assemblages.
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
  layer_spatial(w_asia, fill = NA) +
  layer_spatial(archaeo_sites, size = 0.5) +
  scale_linewidth_identity() +
  coord_sf(crs = w_asia_albers)
```

@@

```{r data-occ}
# Download occurrence data from GBIF, generating a citable derived dataset
if (length(dir_ls(derived_data("gbif"))) == 0) {
  # Requires GBIF credentials as environment variables, see:
  # https://docs.ropensci.org/rgbif/articles/gbif_credentials.html
  message("Reacquiring GBIF occurrences...")
  gbif_download <- occ_download(
    pred_in("taxonKey", flora$gbif_key),
    pred_within(wk::as_wkt(w_eurasia)),
    pred_default(), # Filter common data quality issues, see ?rgbif::pred_default() for details
    # pred_lt("coordinateUncertaintyInMeters", 1000), # too strict for W. Asia!
    format = "SIMPLE_CSV"
  ) |>
    occ_download_wait() # wait for dataset to be prepared

@@

flora_occ_meta <- read_csv(dir_ls(derived_data("gbif"), glob = "*.csv")[1],
                           show_col_types = FALSE)

# Generate GBIF citation
writeLines(paste0(r"(
@misc{GBIFDataDownload,
  title = {GBIF {Occurrence} {Download}},
  author = {{GBIF}},
	publisher = {The Global Biodiversity Information Facility},
  date = {2025-08-08},
  year = {2025},
  url = {https://doi.org/)", flora_occ_meta$doi, r"(}
})"), here("analysis/gbif.bib"))

@@

Georeferenced occurrence data for West Eurasia between 0 and 60° of latitude was obtained from the Global Biodiversity Information Facility (GBIF) using via and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
The GBIF dataset [@GBIFDataDownload] excluded fossil occurrences, recorded absences, and records with missing or dubious coordinates.
Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we did not attempt to model `r english(nrow(filter(flora_all, n_occ < min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
Multiple records of the same taxon at the same coordinate were discarded because they do not impart information to the model.
The resulting cleaned dataset used to train our niche models comprises `r format(flora_occ_meta$totalRecords, scientific = FALSE)` occurrences from `r flora_occ_meta$numberDatasets` constituent datasets.

```{r fig-occ-map, message=FALSE, fig.cap=paste0("Georeferenced occurrence records from West Eurasia used to train models (N=", nrow(unnest(flora, occ)), "). Inset, right: prediction region (West Asia).")}
fig_occ_map <- flora |>
 unnest(occ) |>
 distinct(latitude, longitude) |>
 st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
 ggplot() +
 annotation_spatial(ne_land, fill = "#ffffff") +
 geom_sf(shape = ".") +
 coord_sf(crs = w_asia_albers, default_crs = latlong)

fig_occ_map_inset <- fig_occ_map +
  coord_sf(
    crs = w_asia_albers, default_crs = latlong,
    xlim = w_asia[c(1,3)], ylim = w_asia[c(2,4)], expand = FALSE
  )

fig_occ_map <- fig_occ_map +
 layer_spatial(w_asia, fill = NA)

fig_occ_map + fig_occ_map_inset
```

GBIF is currently the best available general-purpose occurrence dataset for the West Asia region, its coverage is uneven both geographically and from species to species.
The Southern Levant, and Israel specifically, is significantly more densely sampled than other parts of West Asia (@fig-occ-map).

```{r data-background}
if (!file_exists(derived_data("flora_background.Rdata"))) {
  flora <- flora |>
    mutate(
      occ = map(occ, mutate, present = TRUE, weight = 1),
      bg = map(occ, function(x, region, n) { 
        background_sample(
          region = region, 
          N = n,
          coord_x = "longitude", coord_y = "latitude",
          present = FALSE,
          weight = (1 / n) * nrow(x) # weight presences and background equally
        )
      }, region = w_eurasia_land, n = n_background),
      occ = map2(occ, bg, bind_rows)
    ) |>
    select(-bg)
  saveRDS(flora, derived_data("flora_background.Rdata"))
}
flora <- readRDS(derived_data("flora_background.Rdata"))
```

@@

```{r data-climate, warning=FALSE, message=FALSE}
# Reacquire paleoclim data if needed:
if (!file_exists(derived_data("climate.nc"))) {
  if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
    message("Redownloading Paleoclim data...")
    source(here("analysis", "data", "data-paleoclim.R"))
  }

@@

climate <- read_stars(derived_data("climate.nc"), proxy = TRUE, quiet = TRUE)
# Recover dimension values lost in NC conversion
climate <- st_set_dimensions(climate, "period", values = climate_periods$code)
```

@@

```{r data-flora-predictors, warning = FALSE}
bind_extract <- function(x, y) {
  extract <- st_extract(y, st_transform(x, st_crs(y)))
  cbind(x, select(tibble(extract), -geometry))
}

flora <- flora |> 
  mutate(
    occ = map(occ, st_as_sf, coords = c("longitude", "latitude"), crs = latlong, 
              remove = FALSE),
    occ = map(occ, bind_extract, st_as_stars(hack(climate, "period", "cur"))),
    occ = map(occ, bind_extract, st_as_stars(terrain)),
    occ = map(occ, bind_extract, st_as_stars(soil))
  )
```

@@

```{r niche-data}
# Split training and test sets for model assessment 
flora <-  flora |>
  mutate(
    # De-sf
    occ = map(occ, \(x) select(tibble(x), -geometry)),
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE)))),
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
  )

@@

```{r niche-predict, cache.lazy=FALSE}
# NOTE: high memory usage
if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
  flora <- flora |>
    mutate(
      pred = map(niche, collect_predictions),
      palaeodist = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
                       obj = env_w_asia, .progress = TRUE),
      # Binarize
      palaeodist = map(palaeodist, mutate, present = prediction.TRUE. > 0.5),
    )
  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))

  # Write predictions in TIF format
  palaeodist_dir <- dir_create(here("analysis", "palaeodist"))
  walk2(flora$taxon, flora$palaeodist, function(taxon, palaeodist) {
    filename <- paste0(str_replace(taxon, " ", "_"), ".tif")
    write_stars(
      select(palaeodist, "prediction.TRUE."), 
      path(palaeodist_dir, filename)
    )
  })
}
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))

```

@@

We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition (@tbl-results-summary).
Model accuracy (proportion of correctly classified presence and background samples) ranged between `r percent(filter(niche_test_summary, .metric == "accuracy")$min)` and `r percent(filter(niche_test_summary, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "accuracy")$mean)`.
Sensitivity (proportion of correctly classified presence samples) ranged between `r percent(filter(niche_test_summary, .metric == "sens")$min)` and `r percent(filter(niche_test_summary, .metric == "sens")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "sens")$mean)`.
The area under the models' receiver operating characteristic curves (ROC-AUC) was on average `r round(filter(niche_test_summary, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_summary, .metric == "accuracy")$sd, 3)`.
Model sensitivity is loosely correlated with the number of occurrences available for training (@fig-model-sensitivity-vs-n_occ), with the worst-performing models all having less than 1000 recorded occurrences: `r knitr::combine_words(paste0("*", filter(unnest(flora, niche_test_metrics), .metric == "sens" & .estimate < 0.50)$taxon, "*"))`.
Test metrics and ROC curves for the individual models are included in the appendix.

```{r niche-predict-archaeo, cache.lazy=FALSE}
# Extract predictors and make predictions for archaeobotanical occurrences
if (!file.exists(derived_data("archaeo_pred.Rdata"))) {
  archaeo_pred <- flora |>
    unnest(archaeo) |>
    # Add dummy GBIF id
    mutate(gbif_id = 0) |>
    nest(.by = c(climate_period), .key = "archaeo") |>
    mutate(
      archaeo = map(archaeo, st_as_sf, coords = c("longitude", "latitude"),
                    crs = latlong, remove = FALSE),
      archaeo = map2(archaeo, climate_period, function(x, period) {
        bind_extract(x, st_as_stars(hack(climate, "period", as.character(period))))
      }),
      archaeo = map(archaeo, bind_extract, st_as_stars(terrain)),
      archaeo = map(archaeo, bind_extract, st_as_stars(soil))
    ) |>
    unnest(archaeo) |>
    nest(.by = any_of(colnames(flora)), .key = "archaeo") |> # Restore original nesting
    mutate(
      archaeo_pred = map2(niche, archaeo, function(x, y) {
        y <- mutate(y, gbif_key = NA, scientific_name = NA, species = NA)
        augment(extract_workflow(x), new_data = y)
      })
    ) |>
    select(taxon, archaeo_pred)

@@

The ability of the hindcast models to predict the occurrence of specific species at archaeological sites is worse, with just `r percent(p_archaeo_pred_correct)` of presences in archaeobotanical assemblages successfully predicted at a threshold of 0.5.
Model sensitivity (proportion of corrected predicted presenses) in relation to the archaeological data is on average `r round(mean(archaeo_pred_summary$p_predicted, na.rm=TRUE), 2)`±`r round(sd(archaeo_pred_summary$p_predicted, na.rm=TRUE), 2)` (@tbl-results-summary).
A full assessment of the hindcasting performance of the individual models can be found in the appendix.

In summary, the majority of our models perform poorly as simple binary classifiers with a threshold of `P(present)>0.5`.
However, this is an arbitrary threshold; lowering it can improve sensitivity (proportion of occurrences correctly predicted) at the cost of specificity (proportion of non-occurrences incorrectly predicted).
There are several approaches to selecting informative thresholds for binary classification in the ENM literature [@LiuEtAl2005; @LiuEtAl2013; @LiuEtAl2015], with most aiming for an optimum compromise between sensitivity and specificity.
We were unable to find a thresholding technique that significantly improved the hindcast models' performance as binary classifiers.
By definition, binarizing model output also discards information from the underlying probabilistic prediction, so we have opted to present results probabilistically as far as possible.
In principle the archaeological occurrences themselves could be used to 'calibrate' the hindcast models and select an optimum threshold, but there aren't enough archaeological attestations of most of the modelled species for this to be practicable at this time.

@@

results_summary <- flora_all |>
  left_join(flora, join_by(taxon, taxon_aliases, archaeo, n_archaeo)) |>
  left_join(archaeo_pred_summary, join_by(taxon)) |>
  mutate(n_occ = coalesce(n_occ.y, n_occ.x), .before = n_occ.x) |>
  # Extract model metrics
  unnest(c(niche_test_metrics), keep_empty = TRUE) |>
  select(-.estimator, -.config) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  # Excluded samples
  mutate(exclude = (n_occ < min_n_occ) & 
           !taxon %in% c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")) |>
  # Prepare table
  select(taxon, taxon_aliases, n_archaeo, n_occ, exclude, accuracy, roc_auc, 
         sens, p_predicted) |>
  arrange(taxon)

@@

## Reduction in potential niche sizes over the Pleistocene/Holocene boundary

@@

* Archaeological deposits are also time-averaged signals, in some cases over quite large spans of time;
* The chronology attached to our archaeobotanical data is imprecise, e.g. primarily based on point estimates of radiocarbon dates [@Michczynski2007], producing a further averaging effect;
* Archaeological occurences could be the result of human transplanting, range expansion and/or transport, especially considering this is the period where agriculture began to be practiced;
* Most archaeobotanical samples in our archaeological test dataset were small, so individual errors and misidentifications have a correspondingly large impact on quantification of model performance.
